---
title: "This is my Course Project 1 and First Markdown Document!"
output: 
  html_document: 
    keep_md: yes
---

##Loading and Pre-Processing the Data

Read the data from the working directory:
```{r, echo=TRUE}
activityData <- read.csv("./activity.csv")
```

Summarize the data to total steps per day:
```{r, echo=TRUE, message=FALSE}
library(dplyr)
```

```{r, echo=TRUE}
sumActivityDay <- summarize(group_by(activityData,date),
    sumSteps=sum(steps,na.rm=TRUE))
```

##What Is the Mean Total Number of Steps Taken Per Day?

Create a histogram of the total number of steps per day:
```{r, echo=TRUE}
hist(sumActivityDay$sumSteps,
     main="Fitbit Activity Data",
     xlab="Number of Steps Taken in a Day")
```

##What Is the Average Daily Activity Pattern?

First, calculate summary statistics.

```{r, echo=TRUE}
meanSteps <- mean(sumActivityDay$sumSteps, na.rm=TRUE)
```
*The mean number of steps is `r meanSteps`.*

```{r, echo=TRUE}
medianSteps <- median(sumActivityDay$sumSteps, na.rm=TRUE)
```
*The median number of steps is `r medianSteps`.*

Then, summarize the data to average steps per interval:
```{r, echo=TRUE}
sumActivityInterval <- summarize(group_by(activityData,interval),
                          meanSteps=mean(steps, na.rm=TRUE))
```

Create a time series plot of the average numbers of steps per interval:
```{r, echo=TRUE}
plot(meanSteps ~ interval, sumActivityInterval,
     main="Average Number of Steps Per Interval",
     xlab="5 Minute Interval",
     ylab="Average Number of Steps",
     type = "l")
```

Calculate the maximum of the average number of steps across intervals:
```{r, echo=TRUE}
maxSteps <- max(sumActivityInterval$meanSteps)
```
The maximum number of steps is `r maxSteps`.

```{r, echo=TRUE}
maxInterval <- sumActivityInterval[which(sumActivityInterval$meanSteps==maxSteps),"interval"]
```
This maximum occurs in interval number `r maxInterval`.

##Inputing Missing Values

Calculate the total number of missing values in the dataset:
```{r, echo=TRUE}
numNAs <- sum(is.na(activityData$steps))
```
The total number of missing values is `r numNAs`.

Create a new dataset with the missing values filled in with the interval average:
```{r, echo=TRUE}
newActivityData <- arrange(merge(activityData,sumActivityInterval, by = "interval"),date)
newActivityData$steps[is.na(newActivityData$steps)] <-
          newActivityData$meanSteps[is.na(newActivityData$steps)]
```

Re-summarize to the day level:
```{r, echo=TRUE}
newSumActivityDay <- summarize(group_by(newActivityData,date),sumSteps=sum(steps, na.rm=TRUE))
```

Create a histogram of the total number of steps per day:
```{r, echo=TRUE}
hist(newSumActivityDay$sumSteps,
     main="Fitbit Activity Data",
     sub="Missing Values Replaced w/ Interval Average",
     xlab="Number of Steps Taken in a Day")
```

Calculate new summary statistics.


```{r, echo=TRUE}
newMeanSteps <- mean(newSumActivityDay$sumSteps, na.rm=TRUE)
```
*The mean number of steps per day is `r newMeanSteps`.*

```{r, echo=TRUE}
newMedianSteps <- median(newSumActivityDay$sumSteps, na.rm=TRUE)
```
*The median number of steps per day is `r newMedianSteps`.*

The impact of imputing the missing data using the interval average is to increase the mean from `r meanSteps` to `r newMeanSteps` and the median from `r medianSteps` to `r newMedianSteps`.

##Are There Differences in Activity Levels between Weekends and Weekdays?

Create a new factor level in the dataset with the day of the week:
```{r, echo=TRUE}
newActivityData$weekday <- weekdays(as.Date(newActivityData$date))
map <- cbind(c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"),
             c("weekday", "weekday", "weekday", "weekday", "weekday", "weekend", "weekend"))
map <- as.data.frame(map)
colnames(map)=c("weekday", "group")
newActivityData <- arrange(merge(newActivityData, map, by = "weekday"),date)
```

Summarize to weekday/weekend:
```{r, echo=TRUE}
sumActivityGroup <- summarize(group_by(newActivityData, group, interval),
                          newMeanSteps=mean(steps, na.rm=TRUE))
```

Make a panel plot by subset:
```{r, echo=TRUE}
par(mfrow=c(2,1),mar=c(2,4,2,2))
weekend <- subset(sumActivityGroup, group=="weekend")
plot(newMeanSteps ~ interval, weekend, main="Weekends", ylab="Mean # of Steps",
     type="l")
weekday <- subset(sumActivityGroup, group=="weekday")
plot(newMeanSteps ~ interval, weekday, main="Weekdays", ylab="Mean # of Steps",
     type="l")
```

